---
title: "cancensus Demo"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    toc: yes
---

### Info

* Author: Fareeza Khurshed
* Date: 2018-03-04
* This demo requires: cancensus, sf, sp, rgdal, leaflet, tidyverse

*Note that an API key is also required via (censusmapper)[censusmapper.ca] and you need to set the cache location as well. This has been commented out of the code but you will need to include it in your code.*


```{r}
require("cancensus")
require("tidyverse")
require("sf")
require("sp")
require("rgdal")
require("leaflet")

#Set API key - this will store it for future use
#DO NOT MAKE YOUR KEY PUBLIC
options(cancensus.api_key = census_api_key)
options(cancensus.cache_path = census_cache_path)
```

### Background

## Census Data

***

* Census data is collected every 5 years
* In 2011, the census was modified on very short notice to make it a non-mandatory response by the Harper Government
* As a result the quality of the data was compromised, to the point it was renamed from 'census' to the National Household Survey (NHS)
* Because of the concerns with the quality of the data many comparisons are using 2005 data, a ten year gap is being used instead of a five year gap
* The census data has over 2000+ variables that cover a range of topics from age, income, household composition, langauge and ethnicity

## cancensus Package

***

* Covers three census periods, 2005, 2011, 2016

To see the census periods available:

```{r}
list_census_datasets()
```


Census data measures each individual. Because of privacy, the data is aggregated to spatial areas. The spatial breakdown roll up. The levels most commonly used are:

* Province
* CMA - Metropolitan Areas
* DA - Dissemination Area
* CSD - Census Subdivision
* CD - Census Divisions
* CT - Census Tracts
* FED - Federal Electoral Districts
* PED - Provincial Electoral Districs (responsibility of AB Govt)

To see the spatial areas available and the codes that correspond:
```{r}
list_census_regions("CA16")
```


#### To find the name of the region for Edmonton, it's easiest to search for it:
```{r}
list_census_regions("CA16") %>% filter(name == 'Edmonton')
```



## Objective: Examine the median income for households in Alberta
*Use CSD level to limit data (DA is too small)
*Use 2011 census
*Examine the distribution
*See the spatial distribution

#### Obtain the data first:
```{r}
AB <- get_census(dataset='CA16', 
                 regions=list(PR="48"),
                 vectors="v_CA16_2397", 
                 level='CSD', 
                 quiet = TRUE, 
                 geo_format = 'sf', 
                 labels = 'short')

print(AB[, c(5, 7, 14)])
```

#### Check the summary statistics for viewing:
```{r}
summary(AB$v_CA16_2397)
```

#### Create a histogram of the median:
```{r}
ggplot(data=AB, aes(AB$v_CA16_2397)) + 
  geom_histogram(breaks=seq(0, 250000, by = 10000), 
                 col="black", 
                 fill="blue", 
                 alpha = .9) + 
  labs(title="Histogram for Median Income") +
  labs(x="Median Income", y="Count") 
```

#### Basic Spatial Plot, automatic

```{r}
plot(AB[14], main = "AB Median Household Income by CSD")
```

### Interactive Plot via, leaflet

```{r}
#labels for pop-ups
income_labels <- sprintf(
  "<strong>%s</strong><br/>%g Median Household Income",
  AB$name, AB$v_CA16_2397
) %>% lapply(htmltools::HTML)

#bins for colours/legends
bins <- c(0, 10000, 20000, 30000, 40000, 50000, 60000, 70000, 90000, 100000, 125000, 150000, 200000, Inf)
#colour palette selection
pal <- colorBin("RdYlBu", domain = AB$v_CA16_2397, bins = bins)

#create graph
leaflet(AB) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal(v_CA16_2397),
              color = "white",
              weight = 1,
              opacity = 1,
              fillOpacity = 0.65,
              label = income_labels,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>%
  addLegend("bottomright", pal = pal , values = ~v_CA16_2397, opacity = 1, title="Median Household Income by CSD")
```

Unfortunately Edmonton is a single CSD, so this doesn't give us any indication of the distribution within Edmonton. 


### Repeat for only Edmonton, at a DA level

```{r}
edmonton <- get_census(dataset='CA16', regions=list(CMA="48835"), vectors=c("v_CA16_2397"), level='DA', geo_format = "sf", labels='short', quiet=TRUE)


income_labels <- sprintf(
  "<strong>%s</strong><br/>%g Median Income",
  edmonton$GeoUID, edmonton$v_CA16_2397
) %>% lapply(htmltools::HTML)

#names(edmonton)


bins <- c(0, 30000, 60000, 90000,120000, 150000, 200000, 250000, Inf)
pal <- colorBin("RdYlBu", domain = edmonton$v_CA16_2397, bins = bins)
leaflet(edmonton) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~pal(edmonton$v_CA16_2397),
              color = "white",
              weight = 1,
              opacity = 1,
              fillOpacity = 0.65,
              label = income_labels,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>%
  addLegend("bottomright", pal = pal , values = ~edmonton$v_CA16_2397, opacity = 1, title="Average Household Income by DA")
```

Thanks for reading. 

Credit to: @vb_jens, Jens Von Bergman at [Censusmapper](censusmapper.ca) who developed and supports this package. 
